Starcode: an exact algorithm for sequence clustering

### 2014-07-02 doc/document.tex
Corrected a bug. Added plots for the manuscript.

\begin{table}[h]
\centering
\caption{Software execution parameters}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|l|l|}
    \hline
    starcode & \texttt{-t 20}\\ \hline
    CD-HIT 4.6.1 & \texttt{-c 0.925 -T 20 -M 0}\\ \hline
    USEARCH 7.0.1090  & \texttt{-threads 20 -cluster\_fast -id 0.925}\\ \hline
        Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
    Bowtie2 2.1.0 & \texttt{-f -a -p 20 --no-hd --very-sensitive}\\ \hline
        BWA indexer & \texttt{bwa index -a is}\\ \hline
    BWA 0.7.9a & \texttt{mem -t 20 -a -k1 -B0} \\ \hline
    GEM indexer & \texttt{gem-indexer -T 20} \\ \hline
    GEM 1.423 & \texttt{-e3 -s -T 20 --granularity 100000}\\ \hline
    \emph{sequential R} & \texttt{method='lv', maxDist=2}\\ \hline
\end{tabular}
}
\label{tab:soft-options}
\end{table}

@@

We generated 1 million random 40-mers that were duplicated 47 times. For each
40-mer, we inserted 3 addional mutant sequences to the pool by sampling again
3 nucleotides
at random positions. The probability that such a mutant sequence is within a
distance 3 of any other 40-mer is of the order of $10^{-12}$ and can be neglected.
Each cluster thus consists of 50 sequences, of which 3 have a distance at most
3 from the canonical representant.

Using 20 cores, starcode clustered the shuffled 50 million sequences in 4 minutes
without a single error (the output consisted of 1 million clusters of size 50).
We submitted the unique sequences of the dataset to CD-HIT and USEARCH. The
results are summarized in Figure~\ref{fig:04}a and \ref{fig:04}b. The running
times of CD-HIT and USEARCH
were 59 and 64 minutes respectively, and they identified 2,362,352 and 2,739,954
clusters respectively. Both algorithms rely on heuristics, fact that explains why the
number of clusters is off target. These results clearly demonstrate that starcode
outperforms other clustering algorithms in speed and accuracy on this particular task.

>>>>

\begin{table}[h]
\centering
\caption{Software execution parameters}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|l|l|}
    \hline
    starcode & \texttt{-t 20}\\ \hline
    CD-HIT 4.6.1 & \texttt{-c 0.925 -T 20 -M 0}\\ \hline
    USEARCH 7.0.1090  & \texttt{-threads 20 -cluster\_fast -id 0.925}\\ \hline
        Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
    Bowtie2 2.1.0 & \texttt{-f -a -p 20 --no-hd --very-sensitive}\\ \hline
        BWA indexer & \texttt{bwa index -a is}\\ \hline
    BWA 0.7.9a & \texttt{mem -t 20 -a -k1 -B0} \\ \hline
    GEM indexer & \texttt{gem-indexer -T 12} \\ \hline
    GEM 1.423 & \texttt{-e3 -s4 -T12 --granularity 100000}\\ \hline
    \emph{sequential R} & \texttt{method='lv', maxDist=2}\\ \hline
\end{tabular}
}
\label{tab:soft-options}
\end{table}

@@

We generated 1 million random 40-mers that were duplicated 47 times.
For each 40-mer, we inserted 3 addional mutant sequences to the pool
by sampling again 3 nucleotides at random positions. The probability
that such a mutant sequence is within a distance 3 of any other
40-mer is of the order of $10^{-12}$ and can be neglected. Each
cluster thus consists of 50 sequences, of which 3 have a Levenshtein
distance 3 from the canonical representant.

Using 12 cores, starcode clustered the shuffled 50 million sequences
in 4 minutes and 32 seconds without a single error (the output
consisted of 1 million clusters of size 50).  The running times
of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
identified 2,362,352 and 2,739,954 clusters respectively. The results
are summarized in Figure~\ref{fig:04}. Both algorithms rely on
heuristics, explaining why the number of clusters is off target. These
results clearly demonstrate that starcode outperforms other clustering
algorithms in speed and accuracy to cluster random sequences.

We also compared starcode on the biological benchmark dataset proposed
by \cite{pmid23060610}.
